Anomaly detection with kernel density estimation on manifolds

### 2022-05-03 paper/paper.Rmd
Add S2 example reference

---
title: "Anomaly detection with kernel density estimation on manifolds"
author:
- familyname: Cheng
  othernames: Fan
  address: Monash University
  email: Fan.Cheng@monash.edu
  # correspondingauthor: true
- familyname: Hyndman
  othernames: Rob J
  address: Monash University
  email: Rob.Hyndman@monash.edu
- familyname: Panagiotelis
  othernames: Anastasios
  address: University of Sydney
  email: Anastasios.Panagiotelis@sydney.edu.au
abstract: "Kernel density estimation based on random samples often involves a crucial bandwidth selection. For data points lying on the Riemannian manifolds embedded in a high-dimensional space, manifold learning aims to find the low-dimensional representation of the data but there is local distortion for each observation, which failed to be captured by the kernel density estimation. We propose a new kernel density estimation of manifold learning embedding with the Riemannian metric being the pointwise bandwidth matrix and further identify outliers with the lowest densities. The Riemannian metric measures the direction and angle of the local distortion induced by the non-linear functions in manifold learning and is used for smoothing the density estimates. To compare our proposed method with fixed bandwidth KDE, we run two simulations with a 2-D meta data mapped into a 3-D swiss roll or twin peaks shape and a 5-D semi-hypersphere meta data mapped in 100-D space, and show that the proposed estimator could improve the density estimates given a good manifold learning embedding and gives higher correlations between ranks of true density and estimated density. For application, we use the Irish smart meter data and explore the distributions of different households and periods of the week. Both density estimators are compared by examing the distributions of the most typical households with highest densities and the most anomalous households with lowest densities."
keywords: "manifold learning, variable bandwidth, Riemannian metric, highest density region, Gaussian kernels"
# wpnumber: 03/2021
# jelcodes: C55, C65, C80
blind: false
cover: false
toc: false
lot: false
bibliography: references.bib
biblio-style: authoryear-comp
output:
  monash::workingpaper:
    fig_caption: yes
    fig_height: 5
    fig_width: 8
    includes:
      in_header: preamble.tex
      # after_body: appendix.tex
    keep_tex: yes
    number_sections: yes
    citation_package: biblatex
    latex_engine: pdflatex
---

@@

(ref:vkdecaption) The proposed schematic for variable kernel density estimation with recovered geometry.
```{r vkde, fig.align = 'center', out.width = "95%", fig.cap = "(ref:vkdecaption)", echo = FALSE, eval=TRUE}
knitr::include_graphics("figures/vkde.png")
```

@@

\begin{algorithm}[!htb]
  \caption{Learn metric algorithm in \cite{Perrault-Joncas2013-pq} }
  \label{alg:learnmetric}
  \DontPrintSemicolon
  \SetAlgoLined
  \SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}\SetKwInOut{Parameter}{parameter}\SetKwInOut{OptParameter}{optimization parameter}
  \Input{ high-dimensional data $x_i \in \mathcal{R}^s$ for all $i=1,\ldots,N$ }
  \Output{ low-dimensional data $y_i \in \mathcal{R}^d$ and its Riemannian metric $h_i$ for all $i=1,\ldots,N$ }
  \Parameter{ embedding dimension $d$, bandwidth parameter $\sqrt{\varepsilon}$, manifold learning algorithm }
  \OptParameter{ manifold learning parameters }
  \BlankLine
  \begin{algorithmic}[1]

@@

  \end{algorithmic}
\end{algorithm}

# Outlier detection on manifold learning embedding

@@

\begin{algorithm}[!htb]
  \caption{Variable kernel density estimates with Riemannian metric}
  \label{alg:vkderm}
  \DontPrintSemicolon
  \SetAlgoLined
  \SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}\SetKwInOut{Parameter}{parameter}
  \Input{ high-dimensional data $x_i$ for all $i=1,\ldots,N$ }
  \Output{ outliers embedding coordinates $y_1, \dots, y_{n\_outliers}$ with their estimated densities $f_1, \dots, f_{n\_outliers}$ }
  \Parameter{ number of outliers $n\_outliers$, embedding dimension $d$ }
  \BlankLine
  \begin{algorithmic}[1]

@@

  \end{algorithmic}
\end{algorithm}

If the embedding dimension $d=2$, the HDR plots [@Hyndman1996-lk] could be used to identify the relative location of outliers in the embedding and find which observations lie in the highest density region of specified coverage, eg. 1%, 50%, 99%, >99%.

@@

Now we are able to apply different manifold learning algorithms to $(\pmb{X}, \pmb{Y}, \pmb{Z})^\prime$ and reduce the dimension back to $d=2$, and further estimate the density of the 2-D embedding. According to the density estimates, we could rank the data points and then identify which observations lie in the highest density region of specified coverage, eg. 1%, 5%, 50%, 99%, >99%. For each of the five manifold learning methods, namely ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP, \autoref{fig:sroutliers} presents the 2-D embedding plot in the same row, with the colors indicating the densities levels, the left column for true densities from the Gaussian mixture model, the middle column for highest density region plots with densities from our proposed variable KDE method, and the right for similar HDR plots with densities from KDE with fixed bandwidth. The top twenty outliers with the lowest densities are highlighted in black with point indexes in blue. 
From \autoref{fig:metadensity} and the data generating process, we know that there are four highest density regions. However, in all manifold learning embeddings colored with true densities (left column in \autoref{fig:sroutliers} ), except for LLE, the number of highest density regions are not the same as the meta data. When comparing the number of HDRs for variable and fixed bandwidth (middle and right column in \autoref{fig:sroutliers} ), our proposed method with variable bandwidth outperforms fixed bandwidth for ISOMAP, LLE, and Laplacian Eigenmaps (top three rows in \autoref{fig:sroutliers} ). 
In terms of the top 20 outliers found rowwise, variable bandwidth could find most outliers lying on the left area of the embedding in ISOMAP and UMAP, and both methods in LLE embedding could find the outliers in the outer area, but for the other methods, both variable and fixed bandwidth are not detected true outliers accurately. For t-SNE and UMAP embedding, the embedding structure is highly distorted and the points are clustered together in a discontinuous way, which is also shown in the clustered outliers.

@@

In this application, we use the smart meter data from the *CER Smart Metering Project - Electricity Customer Behaviour Trial, 2009-2010* in Ireland [@cer2012-data] between 14 July 2009 and 31 December 2010. The CER dataset^[accessed via the Irish Social Science Data Archive - www.ucd.ie/issda.] records the half-hourly electricity consumption of individual residential and commercial properties, not including energy for cooling or heating systems. We selected the $3,639$ residential data with no missing values during the data collection period for a total of $535$ days.

@@

<!-- Discuss simulation with non-uniform meta data. -->
<!-- The aim of the project is not to evaluate different manifold learning algorithms.  Although ISOMAP worked OK here it is still distorted and in real examples, you will never 'know' that the data are uniform on the sphere.  Instead, the idea of this project is to do kernel density estimation in a way that takes distortion into account.
 -->

# Acknowledgment {-}

@@

\clearpage\appendix
<!-- \section{Appendix: Notions about Riemannian geometry}\label{riemgeo} -->
<!-- %\pagenumbering{arabic} -->
<!-- %\def\thepage{A\arabic{page}} -->

@@

<!-- embedding pushforward metric [@Perrault-Joncas2013-pq] -->                                                                                            
## Exponential map and logarithmic map

@@

## Volumn density function

@@

#### Example: $S^1$

@@

<!-- First, we use the kernel density estimator in \@ref(equ:kde) with a fixed bandwidth using the plug-in bandwidth selector in the R package *ks*.  -->

We use the density estimator in @Pelletier2005-vu. 
<!-- \begin{equation} -->
<!-- \label{eq:pelletier} -->
$$
f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h^d \theta_{X_i}(p)} K(\frac{d_g(p, X_i)}{h}),
$$
<!-- \end{equation} -->
where $d=2$, $h$ is the bandwidth, $\theta_{X_i}(p)$ is the volume density function at sample point $X_i$, and $d_g(p, X_i)$ is the geodesic distance between $p$ and $X_i$ along the manifold $S^1$.

Since the manifold is known ($S^1$), we know the geodesic distance is the arc length between two points on the manifold, which is given by $L(\gamma) = r(\beta - \alpha)$. Therefore, the geodesic distance is
$$
d_g(p, X_i) = \arccos(p) - \arccos(X_i)
$$
As stated in @Pelletier2005-vu, in terms of *geodesic normal coordinates* at $X_i$, the volume density function $\theta_{X_i}(p)$ equals the determinant of the metric $g$ expressed in these coordinates at the logarithm map $\exp_{X_i}^{-1}(p)$, i.e. $|g(\exp_p^{-1}(x_i))|$.

@@

The Riemannian metric $g$ for $S^1$ is given by the inner products on the tangent spaces of $S^1$. For each $X_i \in M$,
$$
g_{X_i}: T_{X_i}M \times T_{X_i}M \rightarrow \pmb{R},
$$
is smooth in a local neightborhood $U$ of $M$.

Then the volume density function $\theta_{X_i}(p)$ is the inner product of the tangent vectors produced by the logarithm map from $X_i$ to $p$, which is given by
$$
\theta_{X_i}(p) = \frac{1}{\sqrt{1 - X_i^2}}.
$$
Therefore, the estimator for $p \in S^1$ is given by
$$
f_N(p) = \frac{1}{N} \sum_{i=1}^{N} |h|
^{-d} \frac{1}{\sqrt{1 - X_i^2}} K(|\arccos(p) - \arccos(X_i)|^{T}h^{-1}|\arccos(p) - \arccos(X_i)|).
$$
The bandwidth $h$ is selected by the plug-in bandwidth selector implemented by the R package *ks*.

>>>>

---
title: "Anomaly detection with kernel density estimation on manifolds"
author:
- familyname: Cheng
  othernames: Fan
  address: Monash University
  email: Fan.Cheng@monash.edu
  # correspondingauthor: true
- familyname: Hyndman
  othernames: Rob J
  address: Monash University
  email: Rob.Hyndman@monash.edu
- familyname: Panagiotelis
  othernames: Anastasios
  address: University of Sydney
  email: Anastasios.Panagiotelis@sydney.edu.au
abstract: "Kernel density estimation based on random samples often involves a crucial bandwidth selection. For data points lying on the Riemannian manifolds embedded in a high-dimensional space, manifold learning aims to find the low-dimensional representation of the data but there is local distortion for each observation, which failed to be captured by the kernel density estimation. We propose a new kernel density estimation of manifold learning embedding with the Riemannian metric being the pointwise bandwidth matrix and further identify outliers with the lowest densities. The Riemannian metric measures the direction and angle of the local distortion induced by the non-linear functions in manifold learning and is used for smoothing the density estimates. To compare our proposed method with fixed bandwidth KDE, we run two simulations with a 2-D meta data mapped into a 3-D swiss roll or twin peaks shape and a 5-D semi-hypersphere meta data mapped in 100-D space, and show that the proposed estimator could improve the density estimates given a good manifold learning embedding and gives higher correlations between ranks of true density and estimated density. For application, we use the Irish smart meter data and explore the distributions of different households and periods of the week. Both density estimators are compared by examing the distributions of the most typical households with highest densities and the most anomalous households with lowest densities."
keywords: "manifold learning, variable bandwidth, Riemannian metric, highest density region, Gaussian kernels"
# wpnumber: 03/2021
# jelcodes: C55, C65, C80
blind: false
cover: false
toc: false
lot: false
bibliography: references.bib
biblio-style: authoryear-comp
output:
  monash::workingpaper:
    fig_caption: yes
    fig_height: 5
    fig_width: 8
    includes:
      in_header: preamble.tex
      # after_body: appendix.tex
    keep_tex: yes
    number_sections: yes
    citation_package: biblatex
    latex_engine: pdflatex
editor_options: 
  markdown: 
    wrap: 72
---

@@

(ref:vkdecaption) The proposed schematic for variable kernel density
estimation with recovered geometry.

```{r vkde, fig.align = 'center', out.width = "95%", fig.cap = "(ref:vkdecaption)", echo = FALSE, eval=TRUE}
knitr::include_graphics("figures/vkde.png")
```

@@

```{=tex}
\begin{algorithm}[!htb]
  \caption{Learn metric algorithm in \cite{Perrault-Joncas2013-pq} }
  \label{alg:learnmetric}
  \DontPrintSemicolon
  \SetAlgoLined
  \SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}\SetKwInOut{Parameter}{parameter}\SetKwInOut{OptParameter}{optimization parameter}
  \Input{ high-dimensional data $x_i \in \mathcal{R}^s$ for all $i=1,\ldots,N$ }
  \Output{ low-dimensional data $y_i \in \mathcal{R}^d$ and its Riemannian metric $h_i$ for all $i=1,\ldots,N$ }
  \Parameter{ embedding dimension $d$, bandwidth parameter $\sqrt{\varepsilon}$, manifold learning algorithm }
  \OptParameter{ manifold learning parameters }
  \BlankLine
  \begin{algorithmic}[1]

@@

  \end{algorithmic}
\end{algorithm}
```
# Outlier detection on manifold learning embedding

@@

```{=tex}
\begin{algorithm}[!htb]
  \caption{Variable kernel density estimates with Riemannian metric}
  \label{alg:vkderm}
  \DontPrintSemicolon
  \SetAlgoLined
  \SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}\SetKwInOut{Parameter}{parameter}
  \Input{ high-dimensional data $x_i$ for all $i=1,\ldots,N$ }
  \Output{ outliers embedding coordinates $y_1, \dots, y_{n\_outliers}$ with their estimated densities $f_1, \dots, f_{n\_outliers}$ }
  \Parameter{ number of outliers $n\_outliers$, embedding dimension $d$ }
  \BlankLine
  \begin{algorithmic}[1]

@@

  \end{algorithmic}
\end{algorithm}
```
If the embedding dimension $d=2$, the HDR plots [@Hyndman1996-lk] could
be used to identify the relative location of outliers in the embedding
and find which observations lie in the highest density region of
specified coverage, eg. 1%, 50%, 99%, \>99%.

@@

Now we are able to apply different manifold learning algorithms to
$(\pmb{X}, \pmb{Y}, \pmb{Z})^\prime$ and reduce the dimension back to
$d=2$, and further estimate the density of the 2-D embedding. According
to the density estimates, we could rank the data points and then
identify which observations lie in the highest density region of
specified coverage, eg. 1%, 5%, 50%, 99%, \>99%. For each of the five
manifold learning methods, namely ISOMAP, LLE, Laplacian Eigenmaps,
t-SNE, and UMAP, \autoref{fig:sroutliers} presents the 2-D embedding
plot in the same row, with the colors indicating the densities levels,
the left column for true densities from the Gaussian mixture model, the
middle column for highest density region plots with densities from our
proposed variable KDE method, and the right for similar HDR plots with
densities from KDE with fixed bandwidth. The top twenty outliers with
the lowest densities are highlighted in black with point indexes in
blue. From \autoref{fig:metadensity} and the data generating process, we
know that there are four highest density regions. However, in all
manifold learning embeddings colored with true densities (left column in
\autoref{fig:sroutliers} ), except for LLE, the number of highest
density regions are not the same as the meta data. When comparing the
number of HDRs for variable and fixed bandwidth (middle and right column
in \autoref{fig:sroutliers} ), our proposed method with variable
bandwidth outperforms fixed bandwidth for ISOMAP, LLE, and Laplacian
Eigenmaps (top three rows in \autoref{fig:sroutliers} ). In terms of the
top 20 outliers found rowwise, variable bandwidth could find most
outliers lying on the left area of the embedding in ISOMAP and UMAP, and
both methods in LLE embedding could find the outliers in the outer area,
but for the other methods, both variable and fixed bandwidth are not
detected true outliers accurately. For t-SNE and UMAP embedding, the
embedding structure is highly distorted and the points are clustered
together in a discontinuous way, which is also shown in the clustered
outliers.

@@

In this application, we use the smart meter data from the *CER Smart
Metering Project - Electricity Customer Behaviour Trial, 2009-2010* in
Ireland [@cer2012-data] between 14 July 2009 and 31 December 2010. The
CER dataset[^2] records the half-hourly electricity consumption of
individual residential and commercial properties, not including energy
for cooling or heating systems. We selected the $3,639$ residential data
with no missing values during the data collection period for a total of
$535$ days.

[^2]: accessed via the Irish Social Science Data Archive -
    www.ucd.ie/issda.

@@

<!-- Discuss simulation with non-uniform meta data. -->

```{=html}
<!-- The aim of the project is not to evaluate different manifold learning algorithms.  Although ISOMAP worked OK here it is still distorted and in real examples, you will never 'know' that the data are uniform on the sphere.  Instead, the idea of this project is to do kernel density estimation in a way that takes distortion into account.
 -->
```
# Acknowledgment {.unnumbered}

@@

\clearpage\appendix

<!-- \section{Appendix: Notions about Riemannian geometry}\label{riemgeo} -->

<!-- %\pagenumbering{arabic} -->

<!-- %\def\thepage{A\arabic{page}} -->

@@

<!-- embedding pushforward metric [@Perrault-Joncas2013-pq] -->

## Exponential map and logarithmic map

@@

## Volume density function

@@

## Example: $S^1$

@@

<!-- First, we use the kernel density estimator in @ref(equ:kde) with a fixed bandwidth using the plug-in bandwidth selector in the R package *ks*.  -->

We use the density estimator in @Pelletier2005-vu. For point
$p \in S^1$, the density at $p$ is estimated by
<!-- \begin{equation} --> <!-- \label{eq:pelletier} --> $$
f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d \theta_{X_i}(p)} K(\frac{d_g(p, X_i)}{r}),
$$ <!-- \end{equation} --> where $d=1$, $h$ is the bandwidth,
$\theta_{X_i}(p)$ is the volume density function at sample point $X_i$,
and $d_g(p, X_i)$ is the geodesic distance between $p$ and $X_i$ along
the manifold $S^1$.

Since the manifold is known ($S^1$), we know the geodesic distance is
the arc length between two points on the manifold, which is given by
$L(\gamma) = r(\beta - \alpha)$ with $r=1$. Therefore, the geodesic
distance is $$
d_g(p, X_i) = \arccos(p) - \arccos(X_i)
$$ As stated in @Pelletier2005-vu, in terms of *geodesic normal
coordinates* at $X_i$, the volume density function $\theta_{X_i}(p)$
equals the determinant of the metric $g$ expressed in these coordinates
at the logarithm map $\exp_{X_i}^{-1}(p)$, i.e. $|g(\exp_p^{-1}(x_i))|$.

@@

The Riemannian metric $g$ for $S^1$ is given by the inner products on
the tangent spaces of $S^1$. For each $X_i \in M$, $$
g_{X_i}: T_{X_i}M \times T_{X_i}M \rightarrow \pmb{R},
$$ is smooth in a local neighborhood $U$ of $M$.

Then the volume density function $\theta_{X_i}(p)$ is the inner product
of the tangent vectors produced by the logarithm map from $X_i$ to $p$,
which is given by $$
\theta_{X_i}(p) = \sqrt{1 - X_i^2}.
$$ Therefore, the estimator for $p \in S^1$ is given by $$
f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r \sqrt{1 - X_i^2}} K\bigg(\frac{\|\arccos(p) - \arccos(X_i))\|}{r}\bigg),
$$ where the bandwidth $r$ is selected by the plug-in bandwidth selector
implemented by the R package *ks*.

### How to derive volume density function $\theta_{X_i}(p)$

Example of $S^2$ in @Henry2009-om.

Let $S^2$ be the two-dimensional sphere of radius 1 and $p \in S^2$. Let
$v$ and $w$ be vectors such that $\|v\| = \|w\| = 1$ and $\{v, w, p\}$
is the orthonormal basis of $\mathbb{R}^3$. Consider the exponential
chart $(U, \Psi)$ induced by the parametrisation
$\Psi^{-1}: B_\pi(0) \rightarrow S^2 \backslash \{-p\}$ given by $$
\Psi^{-1}(s,t) = \cos(r)p + \sin(r)(sv+tw)r^{-1},
$$ if $(s,t) \neq (0,0)$ and $\Psi^{-1}(0,0)=p$ where
$r=\sqrt{s^2 + t^2}$. Note that $r=d_g(p,q)$ if $q = \exp_p(sv+tw)$.
Then we have $$
A(s,t) = \frac{\partial}{\partial \Psi_1} \bigg|_{\Psi^{-1}(s,t)}, \\
B(s,t) = \frac{\partial}{\partial \Psi_2} \bigg|_{\Psi^{-1}(s,t)}.
$$ Then each component of the riemannian metric could be calculated as
$g(A(s,t), A(s,t)), g(A(s,t), B(s,t)), g(B(s,t), B(s,t))$. We obtain
that the volume density on the sphere is $$
\theta_p(q) = \bigg| \det g_q\bigg( \frac{\partial}{\partial \Psi_i}\bigg|_q, \frac{\partial}{\partial \Psi_j}\bigg|_q \bigg) \bigg|^{1/2} \\
= \frac{|\sin(d_g(p,q))|}{d_g(p,g)},
$$ for $q \neq p, -p$, and $\theta_p(p) = 1$.

In $S^1$, polar coordinates: $\phi(\exp_p{vt})^{\prime}(0) = v$

cartesian coordinates:
$\psi(\exp_p{vt})^{\prime}(0) = x_p \cos{v} - \sqrt{1-x_p^2} \sin{v}$
\bigg\_t=0 = -\sqrt{1-x_p^2}v\$
